---
title: "PosterProjectMetIV"
author: "David Jüttner"
date: "`r Sys.Date()`"
output: pdf_document
---

## Setup 

DV: "Thinking about the conflict in Ukraine, to what extent would you support or oppose each of the following? Ukraine fast-tracking European Union membership" (ukraine_memb) --> Ordinal variable ranging from 1 (Strongly support) - 4 (Strongly oppose) --> I recode this so the values are 1 (strongly oppose) - 4 (strongly support)

Methods: OLS (as the baseline model), Lasso, Ridge, Random Forest

The file can take around 15 minutes to run.

```{r - General Setup}

library(tidyverse)
library(glmnet)
library(tidymodels)
library(ranger)
library(knitr)
library(ggplot2)
library(doParallel)
library(tune)

# Set your working directory

setwd()

# Load the data

sie_original <- read.csv("2018_2024_trendfile_dataset.csv")

```

```{r - Data Wrangling}

# I only choose observations and respective variables which are available for the year when my outcome variabe "ukraine_memb" was ask (2022-2024)

sie_working <- sie_original %>% 
  select(ukraine_memb, govrec, starts_with("active_role"), 
         starts_with("country_issues"), starts_with("personal_issues"), 
         referendum, eu_optimism, country_influential, voice_ineu, 
         govtrust20222024, eutrust20222024, euresource_interests, euresource_taxes,
         countrythreat20222024, europeanarmy, crisis_natural_disaster, crisis_debt,
         crisis_unemployment, crisis_refugees, crisis_climate_ch, crisis_pandemic, 
         crisis_military_attack, democsatisfaction_country, democsatisfaction_eu, 
         selfidentity, countrydefence, russia, natoimportance, 
         pastvote, party_name_short_parlgov, family_name_parlgov, left_right_parlgov,
         starts_with("euled"), leftright, ukraine_refugees, economicsituation, 
         areatype, work, age_c, education, gender) %>% 
  filter(ukraine_memb < 5) %>% 
  # Coding the relevant values for NAs
  mutate(govrec = na_if(govrec, 3),
         referendum = na_if(referendum, 4),
         eu_optimism = na_if(eu_optimism, 5),
         country_influential = na_if(country_influential, 6),
         govtrust20222024 = na_if(govtrust20222024, 5),
         eutrust20222024 = na_if(eutrust20222024, 5),
         euresource_taxes = na_if(euresource_taxes, 3),
         countrythreat20222024 = na_if(countrythreat20222024, 7),
         europeanarmy = na_if(europeanarmy, 5),
         crisis_natural_disaster = na_if(crisis_natural_disaster, 3),
         crisis_debt = na_if(crisis_debt, 3),
         crisis_unemployment = na_if(crisis_unemployment, 3),
         crisis_refugees = na_if(crisis_refugees, 3),
         crisis_climate_ch = na_if(crisis_climate_ch, 3),
         crisis_pandemic = na_if(crisis_pandemic, 3), 
         crisis_military_attack = na_if(crisis_military_attack, 3),
         democsatisfaction_country = na_if(democsatisfaction_country, 11),
         democsatisfaction_eu = na_if(democsatisfaction_eu, 11),
         selfidentity = na_if(selfidentity, 6),
         countrydefence = na_if(countrydefence, 4),
         russia = na_if(russia, 4),
         natoimportance = na_if(natoimportance, 5),
         leftright = ifelse(leftright %in% c(8, 100), NA, leftright),
         ukraine_refugees = na_if(ukraine_refugees, 5),
         economicsituation = na_if(economicsituation, 6),
         areatype = na_if(areatype, 6)) 
  

colSums(is.na(sie_working))

# I removed all observations for which >= 15,000 observations have NA (25%)

sie_working <- sie_working %>% 
  select(-starts_with("euled"), -country_influential,
         -pastvote, -party_name_short_parlgov, euresource_taxes) %>% 
  mutate(ukraine_memb = 5 - ukraine_memb)

party_fam_dummies <- model.matrix(~ family_name_parlgov - 1, data = sie_working)
sie_working <- cbind(sie_working, party_fam_dummies)

sie_working <- sie_working %>% 
  select(-family_name_parlgov, -family_name_parlgov.)

sie_working$left_right_parlgov[sie_working$left_right_parlgov == ""] <- NA
sie_working$left_right_parlgov[sie_working$left_right_parlgov == ","] <- NA

sie_working$left_right_parlgov <- as.numeric(as.character(sie_working$left_right_parlgov))

# I decide for listwise deletion which gives me close to 10,000 observations at the end with full data

sie_working <- na.omit(sie_working)

```

```{r - Creating the First-Order-Interactions}

y <- sie_working$ukraine_memb

predictors <- as.data.frame(sie_working[, setdiff(names(sie_working), "ukraine_memb")])

# I created First-Order-Interactions to have more variables to reliably test the different methods

sie_working <- as.data.frame(model.matrix(~ .^2, data = predictors)[, -1])
sie_working$ukraine_memb <- y

```


```{r - Train and Test Data Setup}

set.seed(123)

n <- nrow(sie_working)

# I choose to do a Training/Test split of 80/20.
# This is a bit arbitrary, as other splits would also be possible

train_Index <- sample(n, size = 0.8 * n)

train_data <- sie_working[train_Index, ]
test_data <- sie_working[-train_Index, ]

```

```{r - Parallelization Start}

# Paralellization so my code runs faster (particularly for the Random Forests)

cores <- parallel::detectCores() - 1
cl <- makeCluster(cores)
registerDoParallel(cl)

```


## OLS Baseline

```{r - OLS Baseline Model}

set.seed(123)

# I tried cross-validation (cv) for the OLS baseline just out of interest

K <- 10
n_train <- nrow(train_data)

fold_id <- sample(rep(1:K, length.out = n_train))

cv_mse <- numeric(K)

for (k in 1:K) {
  
  train_data_k <- train_data[fold_id != k, ]
  test_data_k  <- train_data[fold_id == k, ]
  
  model_ols <- lm(ukraine_memb ~ ., data = train_data_k)
  
  pred_k <- predict(model_ols, newdata = test_data_k)
  
  cv_mse[k] <- mean((test_data_k$ukraine_memb - pred_k)^2)
}

model_ols_mse <- mean(cv_mse)
model_ols_mse

model_ols_final <- lm(ukraine_memb ~ ., data = train_data)

pred_test <- predict(model_ols_final, newdata = test_data)

model_ols_test_mse <- round(mean((test_data$ukraine_memb - pred_test)^2), 4)
model_ols_test_mse

# Standard Error

errors_ols <- (test_data$ukraine_memb - pred_test)^2
se_ols <- sd(errors_ols) / sqrt(length(errors_ols))
se_ols

```

## Lasso

```{r - Lasso Model with Lambda min}

set.seed(123)

lasso_train <- model.matrix(ukraine_memb ~ ., data = train_data)
lasso_y_train <- train_data$ukraine_memb

lasso_test  <- model.matrix(ukraine_memb ~ ., data = test_data)
lasso_y_test  <- test_data$ukraine_memb

# Using cross-validation

lasso_cv <- cv.glmnet(lasso_train, lasso_y_train, alpha = 1)

# Setting lambda min. to get the best possible prediction

lasso_minlambda <- predict(lasso_cv, newx = lasso_test, s = "lambda.min")

mse_lasso_minlambda <- round(mean((lasso_y_test - lasso_minlambda)^2), 4)
mse_lasso_minlambda

# Standard Error

se_lasso_minlambda <- sd((lasso_y_test - lasso_minlambda)^2) / sqrt(length(lasso_y_test))
se_lasso_minlambda

```


```{r - Lasso Model with 1SE Lambda}

set.seed(123)

# Using cross-validation

lasso_cv <- cv.glmnet(lasso_train, lasso_y_train, alpha = 1)

# I wanted to try what result lambda 1se gives me

lasso_lambda_one_se <- predict(lasso_cv, newx = lasso_test, s = "lambda.1se")

mse_lasso_lambda_one_se <- round(mean((lasso_y_test - lasso_lambda_one_se)^2), 4)
mse_lasso_lambda_one_se

```

## Ridge

```{r - Ridge Model with Lamdba min}

set.seed(123)

ridge_train <- model.matrix(ukraine_memb ~ ., data = train_data)
ridge_y_train <- train_data$ukraine_memb

ridge_test  <- model.matrix(ukraine_memb ~ ., data = test_data)
ridge_y_test  <- test_data$ukraine_memb

# Using cross-validation

ridge_cv <- cv.glmnet(ridge_train, ridge_y_train, alpha = 0)

# Setting lambda min. for to get the best possible prediction

ridge_minlambda <- predict(ridge_cv, newx = ridge_test, s = "lambda.min")

mse_ridge_minlambda <- round(mean((ridge_y_test - ridge_minlambda)^2), 4)
mse_ridge_minlambda

# Standard Error

se_ridge_minlambda <- sd((ridge_y_test - ridge_minlambda)^2) / sqrt(length(ridge_y_test))
se_ridge_minlambda

```


```{r - Ridge Model with 1SE Lambda}

set.seed(123)

# Using cross-validation

ridge_cv <- cv.glmnet(ridge_train, ridge_y_train, alpha = 0)

# I wanted to try what result lambda 1se gives me

ridge_lambda_one_se <- predict(ridge_cv, newx = ridge_test, s = "lambda.1se")

mse_ridge_lambda_one_se <- round(mean((ridge_y_test - ridge_lambda_one_se)^2), 4)
mse_ridge_lambda_one_se

```

## Random Forest

```{r - Random Forest Model various Specifications}

set.seed(123)

# Defining the folds for cross-validation

folds <- vfold_cv(train_data, v = 10)

# Supporting the parallelization

ctrl <- control_grid(
  allow_par = TRUE,
  parallel_over = "everything"
)

rf_model <- 
  rand_forest(
    trees = tune(),
    min_n = tune()
  ) %>% 
  set_engine("ranger", num.threads = cores) %>% 
  set_mode("regression") # Regression task

rf_wf <- workflow() %>% 
  add_model(rf_model) %>% 
  add_formula(ukraine_memb ~ .) 

rf_tuned <- rf_wf %>%  
  tune_grid(resamples = folds,
            grid=expand_grid(min_n = c(10, 50, 100), # Different parameters for min. N per split
                             trees = c(100, 500, 1000)), # Different parameters for N of trees
            control = ctrl)

rf_tuned %>%  
  show_best(metric="rmse", n = 5)

rf_best <- rf_tuned %>% 
  select_best(metric="rmse")

rf_final <- 
  rf_wf %>% 
  finalize_workflow(rf_best) %>%  
  fit(data = train_data)

rf_pred <- predict(rf_final, test_data) %>%
  bind_cols(test_data)

rf_mse <- round(mean((rf_pred$ukraine_memb - rf_pred$.pred)^2), 4)
rf_mse

# Standard Error

errors_rf <- (rf_pred$ukraine_memb - rf_pred$.pred)^2
se_rf <- sd(errors_rf) / sqrt(length(errors_rf))
se_rf


```

```{r - Parallelization Stop}

stopCluster(cl)

```

## For Latex

```{r - Latex Transformation}

# Horizontal Table

mse_table_horizontal <- data.frame(
  "OLS (Baseline)" = model_ols_test_mse,
  "Lasso Lambda min." = mse_lasso_minlambda,
  "Ridge Lambda min." = mse_ridge_minlambda,
  "Random_Forest" = rf_mse
)

mse_table_horizontal

latex_table_horizontal <- kable(mse_table_horizontal, format = "latex", booktabs = TRUE)

cat(latex_table_horizontal)

# With SEs

mse_table_horizontal_se <- data.frame(
  "OLS (Baseline)" = paste0(model_ols_test_mse, " (", round(se_ols, 6), ")"),
  "Lasso Lambda min." = paste0(mse_lasso_minlambda, " (", round(se_lasso_minlambda, 6), ")"),
  "Ridge Lambda min." = paste0(mse_ridge_minlambda, " (", round(se_ridge_minlambda, 6), ")"),
  "Random_Forest" = paste0(rf_mse, " (", round(se_rf, 6), ")")
)

mse_table_horizontal_se

latex_table_horizontal_se <- kable(
  mse_table_horizontal_se,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE
)

cat(latex_table_horizontal_se)

```


```{r - Visualization Lasso and Ridge MSE with Lambda min (log)}

lasso_data <- data.frame(
  lambda = lasso_cv$lambda,
  mse = lasso_cv$cvm,
  model = "Lasso")

ridge_data <- data.frame(
  lambda = ridge_cv$lambda,
  mse = ridge_cv$cvm,
  model = "Ridge")

combined_lasso_ridge <- rbind(lasso_data, ridge_data)

lasso_lambda_min <- lasso_cv$lambda.min
lasso_mse_min <- min(lasso_cv$cvm)

ridge_lambda_min <- ridge_cv$lambda.min
ridge_mse_min <- min(ridge_cv$cvm)

lasso_ridge_comparison <- ggplot(combined_lasso_ridge, aes(x = log(lambda), y = mse, color = model)) +
  geom_line(linewidth = 1) +
  labs(x = "log(Lambda)", y = "MSE", color = "", title = "Figure 1: Comparison between Lasso and Ridge with min Lambda") +
  geom_vline(xintercept = log(lasso_cv$lambda.min), linetype = "dashed") +
  geom_vline(xintercept = log(ridge_cv$lambda.min), linetype = "solid") +
  geom_hline(yintercept = lasso_mse_min, linetype = "dashed") +
  geom_hline(yintercept = ridge_mse_min, linetype = "solid") +
  annotate("text",
           x = log(lasso_lambda_min),
           y = lasso_mse_min,
           label = paste0("Lambda min: ", round(log(lasso_lambda_min), 2),
                          "\nMSE: ", round(lasso_mse_min, 3)),
           hjust = -0.09, vjust = -2.5, size = 3.5) +
  annotate("text",
           x = log(ridge_lambda_min),
           y = ridge_mse_min,
           label = paste0("Lambda min: ", round(log(ridge_lambda_min), 2),
                          "\nMSE: ", round(ridge_mse_min, 3)),
           hjust = -0.09, vjust = -1.5, size = 3.5) +
  theme_bw() +
  theme(legend.position = "bottom")

lasso_ridge_comparison

ggsave("lasso_ridge_comparison.png",  width = 7, height = 7, plot = lasso_ridge_comparison)

```

```{r - Incuding also Ridge (log)}

# Lasso Lambda min

coef_matrix_lasso <- as.matrix(coef(lasso_cv, s = lasso_cv$lambda))

num_predictors_lasso <- apply(coef_matrix_lasso, 2, function(x) sum(x[-1] != 0))

lasso_num_predictors <- data.frame(
  log_lambda = log(lasso_cv$lambda),
  num_predictors_lasso = num_predictors_lasso
)

lasso_lambda_min <- lasso_cv$lambda.min
lasso_mse_min <- min(lasso_cv$cvm)
lasso_mse_min

coef_min_lambda_lasso <- coef(lasso_cv, s = "lambda.min")
num_predictors_min <- sum(coef_min_lambda_lasso[-1, , drop = TRUE] != 0)

# Lasso Lambda 1se

lasso_lambda_1se <- lasso_cv$lambda.1se

coef_1se_lasso <- coef(lasso_cv, s = "lambda.1se")
num_predictors_1se <- sum(coef_1se_lasso[-1, , drop = TRUE] != 0)

lasso_mse_1se <- lasso_cv$cvm[lasso_cv$lambda == lasso_cv$lambda.1se]
lasso_mse_1se

lasso_min_point <- data.frame(
  log_lambda = log(lasso_lambda_min),
  num_predictors = num_predictors_min
)

lasso_1se_point <- data.frame(
  log_lambda = log(lasso_lambda_1se),
  num_predictors = num_predictors_1se
)

# Plot

lasso_comparison_npredictors <- 
  ggplot(lasso_num_predictors, aes(x = log_lambda, y = num_predictors_lasso)) +
  geom_line(linewidth = 1) +
  geom_point(
  data = lasso_min_point,
  aes(x = log_lambda, y = num_predictors),
  inherit.aes = FALSE,
  color = "black",
  size = 3) +
  geom_point(
  data = lasso_1se_point,
  aes(x = log_lambda, y = num_predictors),
  inherit.aes = FALSE,
  color = "black",
  size = 3) +
  geom_vline(xintercept = log(lasso_lambda_min), linetype = "dashed", color = "black") +
  geom_vline(xintercept = log(lasso_lambda_1se), linetype = "dashed", color = "black") +
  annotate("text",
           x = log(lasso_lambda_min),
           y = num_predictors_min,
           label = paste0(
             "Lambda min\nN Predictors: ", num_predictors_min,
             "\nMSE: ", round(lasso_mse_min, 3)),
           hjust = 1.1,
           vjust = 1.25,
           size = 3.5,
           color = "black") +
  annotate("text",
           x = log(lasso_lambda_1se),
           y = num_predictors_1se,
           label = paste0(
             "Lambda 1se\nN Predictors: ", num_predictors_1se,
             "\nMSE: ", round(lasso_mse_1se, 3)),
           hjust = -0.1,
           vjust = -0.25,
           size = 3.5,
           color = "black") +
  labs(x = "log(Lambda)", y = "Number of Selected Predictors", title = "Figure 2: Predictor Selection in Lasso") +
  theme_bw()

lasso_comparison_npredictors

ggsave("lasso_comparison_npredictors.png",  width = 7, height = 7, plot = lasso_comparison_npredictors)


```

```{r - Tree-based Regression Visualization}

rf_results_plot <- rf_tuned %>%
  collect_metrics() %>% 
  filter(.metric == "rmse")  %>% 
  mutate(mse = mean^2) %>% 
  arrange(mse)

random_forests <- ggplot(rf_results_plot, aes(
  x = factor(min_n),
  y = mse, fill = factor(trees))) +
  geom_col(position = "dodge") +
  coord_cartesian(ylim = c(0.7, 0.74)) +
  scale_fill_manual(values = c(
  "100" = "darkblue",
  "500" = "orange",
  "1000" = "darkgreen")) +
  scale_x_discrete(labels = function(x) {
    gsub("\\.", " / ", x)
  }) +
  labs(
    x = "Min N per Split",
    y = "MSE", fill = "N of Trees", title = "Figure 3: Different Combinations of Random Forests") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

random_forests

ggsave("random_forests.png", width = 7, height = 7, plot = random_forests)

```


